Grove: A Bidirectionally Typed Collaborative Structure Editor Calculus

### 2024-07-11 formalism.tex

A \emph{source} is a pair $(v, p) \in \S = \V \times \P$, where
  vertices are drawn from $\V$,
  and positions are drawn from $\P$.
The \emph{outedges} of a source are the edges that have the source.

An \emph{edge} $\e{=}(w, (v, p), v^\prime) \in \E = \W \times \S \times \V$ represents a directed multi-edge 
    identified by $w \in \W$, 
    originating from source $(v, p) \in \S$, 
    with destination vertex $v' \in \V$.

@@

Given a source language parameterized by a set of constructors $k \in \K$ and an arity function $\K \rightarrow \mathbb{N}$, we define an augmented term language as specified in \autoref{fig:Syntax}. Terms are extended with cases for multi-location references and unicycle references. Additionally, since any position of any constructor may have zero, one, or multiple outgoing edges, we define a secondary sort \textit{SubTerm} to represent these possibilities. A source with no outgoing edges corresponds to an empty hole, a source with one outgoing edge corresponds to an ordinary subterm, and a source with multiple outgoing edges corresponds to a local conflict between multiple terms.

% TODO: Should we be more formal and say "is annotated with" instead of "carries"?
These terms also carry information to support additional operations. To enable the reconstruction of the original graph, each constructor carries the UID of its original vertex, each reference carries the vertex that it refers to, and each subterm carries the UID of the edge that leads to it. To support type hole inference (discussion in Sec. TODO), each empty hole and local conflict carries its source. Finally, to support user navigation of the grove, each reference also carries the UID of the edge that leads to it.

@@

\begin{align}
    \text{parents}_{G}(v) &= \SetOf{v' \in \V \SuchThat{(w, (v', p), v) \in G \land w \in \W \land p \in \P}}
\end{align}

@@

The decomposition of a vertex is specified intuitively as the term obtained by traversing the descendants of the vertex until a leaf or a root vertex of some kind is reached, which not further decomposed, but is denoted in the term as the appropriate reference. The decomposition of a graph is simply the set obtained by decomposing each vertex classified as a root.

\begin{definition}
    $\text{decomp}_{G}(v) = $
\end{definition}

The recomposition of a term is a set of edges, and is specified intuitively as the set of parent-child edges in the term. Terms store vertices on constructors and edge UIDs for each child to enable the completeness of this reconstruction.

\begin{theorem}
    (Recomposability) TODO
\end{theorem}

@@

$\Reposition{v}{p}$ combines edge construction and deletion into a single \emph{atomic} operation. If the cursor is not an empty hole, the corresponding edge(s), as indicated above, are simultaneously deleted and then reconstructed as child(ren) with edge source $(v,p)$.

>>>>

A \emph{location} $\ell = (v, p) \in \L = \V \times \P$ represents the origin of an edge, where vertices are drawn from $\V$ and positions are drawn from $\P$.

An \emph{edge} $\e{=}(w, \ell, v^\prime) \in \E = \W \times \L \times \V$ represents a directed multi-edge 
    identified by $w \in \W$, 
    originating from location $\ell \in \L$, 
    with destination vertex $v' \in \V$.

@@

Given a source language parameterized by a set of constructors $k \in \K$, each associated with a fintie set of positions $arity(k)\subseteq \P$, we define an augmented term language as specified in \autoref{fig:Syntax}. Terms are extended with cases for multi-location references and unicycle references. Additionally, since any position of any constructor may have zero, one, or multiple outgoing edges, we define a secondary sort \textit{ChildTerm} to represent these possibilities. A location with no outgoing edges corresponds to an empty hole, a location with one outgoing edge corresponds to an ordinary subterm, and a location with multiple outgoing edges corresponds to a local conflict between multiple terms.

% TODO: Should we be more formal and say "is annotated with" instead of "carries"?
These terms also carry information to support additional operations. To enable the reconstruction of the original graph, each constructor carries the UID of its original vertex, each reference carries the vertex that it refers to, and each subterm carries the UID of the edge that leads to it. To support type hole inference (discussion in Sec. TODO), each empty hole and local conflict carries its location. Finally, to support user navigation of the grove, each reference also carries the UID of the edge that leads to it.

@@

\begin{align*}
    \text{parents}_{G}(v) &= \SetOf{v' \in \V \SuchThat{(w, (v', p), v) \in G \land w \in \W \land p \in \P}}\\
    \text{children}_{G}(\ell) &= \SetOf{(w,v) \in \W\times\V \SuchThat{(w, \ell, v) \in G}}
\end{align*}

@@

\begin{definition}
    $\text{root}_G(v)$ if $v$ is an NP root, an MP root, or a U root in $G$.
\end{definition}

The decomposition of a vertex is specified intuitively as the term obtained by traversing the descendants of the vertex until a leaf or a root vertex of some kind is reached, which not further decomposed, but is denoted in the term as the appropriate reference. The decomposition of a graph is simply the set obtained by decomposing each vertex classified as a root, provided the vertex actually appears as the origin of an edge in $G$.

\begin{definition}
\begin{align*}
    \text{decompChild}_G(w, v)  &= \begin{cases}
        \multiref  &\quad v\text{ is an MP root in } G\\
        \uniref  &\quad v\text{ is a U root in} G\\
        \text{decompVertex}_{G}(v) &\quad v\text{ is not a root in} G
    \end{cases} 
\end{align*}
\end{definition}

\begin{definition}
\begin{align*}
    \text{decompLocation}_{G}(\ell) &= \begin{cases}
        \ehole &\quad \text{children}_{G}(\ell) = \emptyset \\ \eid{w}{}{\text{decompChild}_G(w, v')} &\quad \text{children}_{G}(\ell) = \SetOf{(w,v')}\\ \conflictHoleTerm{\ell}{\eid{w}{}{\text{decompChild}_G(w, v')} \SuchThat{(w,v') \in \text{children}_{G}(\ell)}} &\quad \abs{\text{children}_{G}(\ell)} > 1 \\
    \end{cases}
\end{align*}
\end{definition}

\begin{definition}
    $\text{decompVertex}_{G}(v = (u,k)) = \genericTerm{u}{\text{decompLocation}_{G}((v,p))}$
\end{definition}

\begin{definition}
    \text{vertices}(G) = \SetOf{v \in \V \SuchThat{(w, (v,p), v') \in G \land w \in \W \land v' \in \V \land p \in \P}}
\end{definition}

\begin{definition}
    $\text{decomp}(G) = \SetOf{\text{decompVertex}_{G}(v) \SuchThat{\text{root}_G(v)} \land v \in \text{vertices}(G)}$
\end{definition}

The recomposition of a term is a set of edges, and is specified intuitively as the set of parent-child edges in the term. Terms store vertices on constructors and edge UIDs for each child to enable the completeness of this reconstruction. The recomposition of a set of terms is the union of the recompositions of the terms.

\begin{definition}
    \begin{align*}
    \text{vertexOfTerm}(\genericTerm{u}{\subtermMV_p}) &= (u,k)\\
    \text{vertexOfTerm}(\multiref) &= v\\
    \text{vertexOfTerm}(\uniref) &= v
    \end{align*}
\end{definition}

\begin{definition}
    \begin{align*}
    \text{recompChildTerm}(\ell, \ehole) &= \emptyset\\
    \text{recompChildTerm}(\ell, \eid{w}{}\termMV) &= \SetOf{(w, \ell, \text{vertexOfTerm}(\termMV))}\cup \text{recompTerm}(t)\\
    \text{recompChildTerm}(\ell, \conflictHoleTerm{\ell}{\eid{w_i}{}\termMV_i}_{i<n}) &= \bigcup_{i<n} (\SetOf{(w_i, \ell, \text{vertexOfTerm}(\termMV_i))}\cup \text{recompTerm}(t_i))
    \end{align*}
\end{definition}

\begin{definition}
    \begin{align*}
        \text{recompTerm}(\genericTerm{u}{\subtermMV_p}) &= \bigcup_{p\in arity(k)}\text{recompChildTerm}(((u,k),p), \subtermMV_p)\\
        \text{recompTerm}(\multiref) &= \emptyset\\
        \text{recompTerm}(\uniref) &= \emptyset
    \end{align*}
\end{definition}

\begin{definition}
    $\text{recomp}(T) = \SetOf{\text{recompTerm}(\termMV) \SuchThat{\termMV \in T}}$
\end{definition}

\begin{theorem}
    (Recomposability) For all subgraphs $G$, $\text{recomp}(\text{decomp}(G))=G$.
\end{theorem}

\subsubsection{Grove Construction}

In order to facilitate user interaction with the term set that results from decomposition, it is refined into a \textit{grove}. A grove is a partition of the term set into three classes, corresponding to the the three kinds of roots: NP, MP, and U. A grove also designates a distinguished child term as the primary component of the program.

Formally, a grove $\gamma$ is a quadruple $(\subtermMV_r, NP, MP, U)$ where $\subtermMV \in ChildTerm$ and $NP, MP, U \subseteq Term$. The construction of a grove from the term set of $G$ requires the designation of a location $(v_r,p_r)$ as the \textit{distinguished root} such that $v_r$ is an NP root in $G$. In that case the grove is defined as follows:

\begin{definition}
\begin{align*}
    \text{distinguishedTerm}(\ell_r=((u_r,k_r),p_r), \Theta) &= \subtermMV_r &\quad \text{ if }
    \genericConstructorTerm{u_r}{k_r}{\subtermMV_r}\in \Theta \\
    \text{distinguishedTerm}(\ell_r=((u_r,k_r),p_r), \Theta) &= \emptyHole{\ell_r} &\quad \text{ otherwise}
\end{align*}
\end{definition}

@@

$\Reposition{v}{p}$ combines edge construction and deletion into a single \emph{atomic} operation. If the cursor is not an empty hole, the corresponding edge(s), as indicated above, are simultaneously deleted and then reconstructed as child(ren) with edge location $(v,p)$.

### 2024-07-11 grove-by-example.tex

Although the collaborative edits discussed so far merge cleanly,
they can also lead to graphs that do not map cleanly to a syntax tree.
We call these states \emph{conflicts}
and consider separate cases for \emph{multi-child}, \emph{multi-parent}, and \emph{unicycle} conflicts.
As with merge conflicts in version-control systems such as git, these all require user intervention to resolve.
We consider common merge scenarios first, then move onto more unusual scenarios.

@@

Suppose Alice and Bob both have the edit state \texttt{w * v + \hole} in \autoref{fig:NestedParts:c}.
Alice moves the cursor to the hole and constructs~\vMultiChildAlice{} as the~\texttt{R} child of~\vWrapPlus{}.
At the same time, Bob constructs~\vMultiChildBob{} in the same position.
Now Alice and Bob have the graphs in \autoref{fig:MultiChild:a} and \autoref{fig:MultiChild:b}, respectively.
Note that these graphs have two different edges,
\eMultiChildAlice{} and \eMultiChildBob{}, for the~\texttt{R} position of~\vWrapPlus{}.

We merge these edits by including \emph{both} \eMultiChildAlice{} and \eMultiChildBob{}
as edges originating from the \texttt{R} position of \vWrapPlus{}.
This results in a multi-child conflict in the~\texttt{R} position of~\vWrapPlus,
shown in \autoref{fig:MultiChild:c},
where there are multiple edges originating from the same position of a vertex.
In the user interface, we flag this conflict with the notation~$\conflictHole{x,y}$.

@@

Such conflicts can be resolved by deleting one of the conflicting edges
and editing the remaining term into the correctly merged value.
For example, Alice could delete \eMultiChildBob{} and then change~\vMultiChildAlice{} into~\texttt{x * y}.
% To ensure users can continue editing while in a conflicted state,
% all conflicts must be resolved before a program can be run.\footnote{It
%   may be possible to develop evaluation models that allow these sorts of conflicts,
%   but that is beyond the scope of this paper.})
% % TODO: need example of fully-automatic resolution and semi-automatic resolution.
% Note that as a convenience to the user, certain simple conflicts might be automatically resolved,
% but we consider this a higher-level, user-interface consideration.
Handling conflicts in this way is similar to handling traditional merge conflicts
in that it presents alternate versions of the conflicted state
that must be transformed by users into the desired result.
Our approach is different in that merge conflicts in a line-based model
may break semantic editor services, such as semantic search or type-preserving refactorings,
that could continue to work inside and outside the conflict in a graph-based model.
% It is just the where the multi-child happens that editor services may break down.
\autoref{sec:Related Work} discusses more differences.

>>>>

The collaborative edits discussed so far merge cleanly,
but in general, merging patches can lead to graphs that do not map cleanly to a conventional syntax tree. We identify several different motifs that might arise, all of which give rise to different kinds of conflicts in the graph decomposition: \emph{local conflicts}, \emph{multi-location conflicts}, and \emph{cyclic location conflicts}. 
As with merge conflicts in version-control systems such as git, these all require user intervention to resolve.

@@

Suppose Alice and Bob both start with the edit state \texttt{w * v + \hole} from \autoref{fig:NestedParts:c}.
Alice moves the cursor to the hole and constructs~\vMultiChildAlice{} as the~\texttt{R} child of~\vWrapPlus{}.
At the same time, Bob constructs~\vMultiChildBob{} at the same location.
Now Alice and Bob have the graphs in \autoref{fig:MultiChild:a} and \autoref{fig:MultiChild:b}, respectively.

When these patches are merged, \emph{both} \eMultiChildAlice{} and \eMultiChildBob{}
appear in the merged graph. 
When decomposing this graph to a syntax tree, 
we resolve this conflict in the~\texttt{R} position of~\vWrapPlus~by decomposing to a \emph{local conflict}, \figureCode{$\binaryConflictHole{x}{y}$}, as 
shown in \autoref{fig:MultiChild:c}.

@@

Such conflicts can be resolved simply by deleting relocating one of the conflicting terms (and editing the remaining term into the correctly merged value, if needed), which would remove the corresponding edge.
For example, Alice could resolve the problem by wrapping \li{x} and \li{y} with a multiplication, effectively moving them to non-conflicting locations.
No special edit actions are needed for conflict resolution.

% To ensure users can continue editing while in a conflicted state,
% all conflicts must be resolved before a program can be run.\footnote{It
%   may be possible to develop evaluation models that allow these sorts of conflicts,
%   but that is beyond the scope of this paper.})
% % TODO: need example of fully-automatic resolution and semi-automatic resolution.
% Note that as a convenience to the user, certain simple conflicts might be automatically resolved,
% but we consider this a higher-level, user-interface consideration.

It is worth noting one special situation: if Alice and Bob both independently filled the hole with structurally identical terms, e.g. \li{x}, Grove would \emph{still} formally identify a conflict, because each \li{x} has a distinct UID. In this situation, it would be reasonable for the system to resolve the conflict without further coordination by deterministically choosing one of the two UIDs, e.g. the smallest one.

When the conflicted terms are similar up to UID differences but not identical, it might be helpful  to give the developer the option to ``push down'' the conflicts as deeply as possible, using a tree differencing algorithm. However, this would increase the number of conflicts overall, so it may not always be preferable. Tree differencing is not fundamental to the collaboration model of Grove.

As we will return to in \autoref{sec:Type System}\todo{sec}, 
a local conflict need not disable type-based editor services (the \textbf{semantic gap problem}). Instead, the system can type check around the conflict (and if the conflicted terms have a common type, the system can infer that type for the location conflict itself).
